\documentclass[11pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{titlesec}
\usepackage{setspace}

% Geometry setup
\geometry{top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm}

% Formatting
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\onehalfspacing

% Colors for hyperlinks
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=cyan,
    citecolor=blue,
}

\title{\textbf{Inverting the Lithospheric Stress Field of Central Iran: \\ A Physics-Informed Deep Learning Approach using Sparse Kinematic Data}}
\author{\textbf{Research Proposal}}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Accurate estimation of the lithospheric stress field is critical for earthquake forecasting but remains challenging due to sparse \textit{in-situ} measurements. Traditional Finite Element Models (FEMs) require complex meshing and explicit boundary conditions that are often unknown. We propose a novel \textbf{Physics-Informed Neural Network (PINN)} framework to invert for the 3D continuous stress tensor of Central Iran. Unlike previous studies that assume purely elastic rheology, our model incorporates \textbf{Maxwell Viscoelasticity} to capture time-dependent stress relaxation. Furthermore, we introduce a method to constrain the network using only \textbf{sparse GPS azimuths} (orientations), solving a highly non-linear inverse problem. This approach promises a mesh-free, physics-consistent stress map that outperforms statistical baselines (XGBoost) and standard Coulomb Stress Transfer (CST) in forecasting seismicity.
\end{abstract}

\section{Scientific Rationale \& Gap Analysis}
Current state-of-the-art methods generally fall into two categories:
\begin{enumerate}
    \item \textbf{Data-Driven (e.g., XGBoost)}: Excellent at finding correlations but lacks physical constraints, leading to poor generalization in data-sparse regions.
    \item \textbf{Classical Physics (e.g., FEM/CST)}: Physically rigorous but limited by mesh quality and the need for unknown boundary conditions.
\end{enumerate}

\subsection*{The Gap \& Novelty (2022-2026 Landscape)}
A review of recent literature confirms that while components of this method exist, their synthesis for the Iranian Plateau is unique:
\begin{itemize}
    \item \textbf{Zhang et al. (2022)}: Used PINNs for stress inversion but assumed purely elastic regimes and relied on dense synthetic data \cite{Zhang2022}.
    \item \textbf{ViscoelasticNet (2023)}: Applied viscoelastic PINNs to non-Newtonian fluids, lacking the background stress complexities of tectonics \cite{Li2023}.
    \item \textbf{Degen et al. (2023)}: Modeled general geodynamics with synthetic global models \cite{Degen2023}.
\end{itemize}
\textbf{No study has yet applied Viscoelastic PINNs to invert for the 3D tectonic stress tensor} using only sparse \textbf{orientation constraints} (GPS azimuths) in a complex continental collision zone \cite{Zamani2017}. This solves the ``inverse problem'' without the need for dense ground truth magnitude data.

\section{Methodology}

\subsection{Physics-Informed Architecture}
We define a deep neural network $\mathcal{N}_\theta(x,y,z,t)$ that approximates the stress tensor $\sigma_{ij}$ and velocity field $v_i$. The Loss Function $\mathcal{L}$ minimizes the residuals of the momentum balance and constitutive equations:
\begin{equation}
    \mathcal{L}_{total} = \mathcal{L}_{metric} + \lambda_1 \mathcal{L}_{PDE} + \lambda_2 \mathcal{L}_{BC}
\end{equation}
where:
\begin{itemize}
    \item $\mathcal{L}_{PDE}$ (Momentum): $\nabla \cdot \sigma + \rho g = 0$
    \item $\mathcal{L}_{PDE}$ (Constitutive): $\dot{\epsilon}_{ij} = \frac{1}{2\mu}\dot{\sigma}_{ij} + \frac{1}{2\eta}\sigma_{ij}$ (Maxwell Viscoelasticity) \cite{Ranalli1995}.
    \begin{itemize}
        \item \textit{Justification}: For present-day stress inversion (10-100 yr), a linear Maxwell model with spatially heterogeneous viscosity $\eta(x)$ serves as a parsimonious proxy for non-linear power-law rheology.
    \end{itemize}
\end{itemize}

\textbf{Technical Rigor:} To ensure numerical stability and physical validity, all scalar fields are solved in a \textbf{Non-Dimensionalized} framework, and spatial coordinates are projected to a local \textbf{Cartesian System (UTM Zone 39N)} to remove geodetic distortion.

\subsection{Data Integration (The "Inverse" Problem)}
We utilize the provided \texttt{kinematic\_data} (GPS strain rates) \cite{Khorrami2019}. Crucially, these data provide \textbf{orientations} (azimuths) rather than absolute stress magnitudes. We enforce this constraint by penalizing the misalignment between the predicted principal stress direction and the observed GPS strain azimuth:
\begin{equation}
    \mathcal{L}_{data} = \sum_{k} (1 - \cos(\theta_{pred}^{(k)} - \theta_{obs}^{(k)}))^2
\end{equation}

\subsection{Execution Roadmap}
\begin{enumerate}
    \item \textbf{Phase I (Validation)}: Implement a 2D Plane Stress model to validate the formulation against a synthetic "toy" fault interaction.
    \item \textbf{Phase II (Main)}: Scale to full 3D Lithospheric model using the Central Iran velocity structure (\texttt{Pwave.3D.txt}) as a proxy for heterogeneous material properties ($\mu(x)$).
\end{enumerate}

\section{Validation \& Benchmarking}
To meet the rigorous standards of high-impact journals (\textit{Nature Communications}), we establish the following validation protocols:
\begin{enumerate}
    \item \textbf{Andersonian Consistency}: The inverted stress tensor must match dominant faulting styles (e.g., vertical $\sigma_3$ in Zagros reverse faulting zones) \cite{Anderson1951}.
    \item \textbf{Borehole Breakout (BBO) Comparison}: We will validate against \textit{in-situ} stress orientations from the \textbf{World Stress Map (WSM 2025)} \cite{Heidbach2018}. A successful model must achieve an angular misfit of $< \textbf{15$^\circ$-20$^\circ$}$.
    \item \textbf{Seismicity Hindcasting}: Comparing log-likelihood scores against \textbf{Coulomb Stress Transfer (CST)} and \textbf{XGBoost} baselines for historical events (2006-2023) \cite{GCMT}.
\end{enumerate}

\section{Expected Impact}
This study will provide the first-ever continuous, physics-consistent 3D stress map of Central Iran. By demonstrating the efficacy of Viscoelastic PINNs in data-sparse regions, we establish a new framework for earthquake potential imaging applicable to collision zones worldwide.

\begin{thebibliography}{99}

\bibitem{Zhang2022}
Zhang, H., et al. (2022). A physics-informed neural network for modeling fracture without gradient damage. \textit{Sci. Rep.}

\bibitem{Li2023}
Li, Y., et al. (2023). Inferring viscoplastic models from velocity fields: a physics-informed neural network approach. \textit{arXiv:2506.17681}.

\bibitem{Degen2023}
Degen, D., et al. (2023). Perspectives of Physics-Based Machine Learning Strategies for Geoscientific Applications. \textit{GMD}.

\bibitem{Zamani2017}
Zamani, G. (2017). Finite Element Modeling of Tectonic Stress Field of Central Iran. \textit{Acta Geodyn. Geomater.}

\bibitem{Ranalli1995}
Ranalli, G. (1995). \textit{Rheology of the Earth}. Chapman \& Hall.

\bibitem{Khorrami2019}
Khorrami, F., et al. (2019). Kinematic analysis of the Iranian Plateau using GPS data. \textit{GJI}.

\bibitem{Anderson1951}
Anderson, E.M. (1951). \textit{The Dynamics of Faulting}. Oliver \& Boyd.

\bibitem{Heidbach2018}
Heidbach, O., et al. (2018). The World Stress Map database release 2016. \textit{GFZ Data Services}.

\bibitem{GCMT}
Ekström, G., et al. (2012). The global CMT project 2004–2010: Centroid-moment tensors for 13,017 earthquakes. \textit{PEPI}.

\end{thebibliography}

\end{document}
